### 2025-10-25 thesis/implementation/index.tex
Self refinement

To extract possible candidates all generated responses are post-processed
through a markdown parser (\texttt{pulldown-cmark}) which parses the model
output and looks for the code-block fence characters \texttt{\`\`\` }.

Subsequently all candidates are processed in reversed order and the first full
valid query is returned as potential candidate. The reversing of processing
order is needed as the model outputs it's thoughts top-to-bottom as a markdown
document with the most likely answer usually being output at the end. This
approach ensures that a returned query candidates are executable and
ensured to contain valid SQL. Thus \textsc{Natural} can aid the
difficulty of generating perfect SQL queries by acknowledging
limitations from large language models and implementing recovery and refinement
mechanisms in the subsequent pipeline flow.

>>>>

To extract possible candidates all generated responses are post-processed
through a markdown parser (\texttt{pulldown-cmark}) which parses the model
output and looks for the code-block fence characters \texttt{```}.
Subsequently all candidates are processed in reversed order and the first full
valid query is returned as potential candidate. The reversing of processing
order is needed as the model outputs it's thoughts top-to-bottom as a markdown
document with the most likely answer usually being output at the end. This
approach ensures that a returned query candidates are executable and
ensured to contain valid SQL. Thus \textsc{Natural} can aid the
difficulty of generating perfect SQL queries by acknowledging
limitations from large language models and implementing recovery and refinement
mechanisms in the subsequent pipeline flow.

\subsubsection{Self-Refinement Mechanism ($\rho$)}

The self-refinement algorithm described in~\ref{design:refinement-function}
is implemented in \texttt{natural-inference::pipeline::refinement}.
This implementation corresponds to the $\rho$ function, providing
automated error correction through execution feedback and iterative
improvement of generated SQL queries.

\paragraph{Error Correction Through Execution Feedback}

The \texttt{Refinement::optimize} method takes a \texttt{RawQueryCandidate} and
attempts to improve it through targeted prompting. The refinement prompt
explicitly instructs the model to ``spot any errors in the SQL query, correct
them'' and provides the original question, schema context, and the candidate
query that needs improvement.

Contrary to the ICL generation which uses few-shot examples, the
refinement process focuses on single-query self-correction with explicit
error-fixing instructions.

\paragraph{Prompt Engineering for Correction}

The refinement prompt uses a structured format similar to ICL generation but
with key differences: it includes the original question as context, provides
the candidate query that needs fixing, and uses explicit correction language.

\begin{verbatim}
    SHOW PROMPT LAYOUT
\end{verbatim}

\paragraph{Model Resource Management}

As both the ICL implementation and the refinement are ultimately generating
SQL, they reuse the same underlying \texttt{SqlModel} from the shared pipeline
context rather than loading separate models. This design choice
significantly reduces memory requirements but potentially impacts refinement
quality compared to having dedicated refinement models with different prompt
conditioning. Due to the limited available hardware the effects of having a
separate, distinc refinement model could not be verified. \citeauthor{XiYan}
have implemented multi model generation pipelines in \citeyear{XiYan} and
achieved promising results which.

Similar to ICL generation, refined responses are processed through the same
\texttt{Markdown} parser to extract the predicted SQL. The system validates each
extracted candidate using \texttt{QueryCandidate::try_from} and returns the first
syntactically valid refinement starting from the bottom. This ensures that
refinement produces executable SQL whilst falling back gracefully if refinement
fails.

\paragraph{Integration with Pipeline}

The refinement module is integrated into the main pipeline after the ICL
generator. It processes the \texttt{RawQueryCandidate} and provides a
self-correction mechanism that improves overall pipeline robustness. Notably both
the unrefined and the refined query candidates are kept in the candidate set
for majority voting.
